Predicting Errors with Second Language Acquisition Modeling

### 2021-06-14 paper.Rmd
Update notebook

```{r, echo=FALSE}
library(tidyverse)
library(patchwork)
library(psych)
```

@@

``` {r, echo=FALSE, results='markup'}
cols_md = c('user_id', 'country', 'days', 'client',
            'session', 'format', 'time', 'session_id',
            'n_tokens', 'n_errors', 'prompt')
col_types_md = cols('f','f','d','f','f','f','i','f','i', 'i', 'f')

@@

```{r, echo=FALSE}
# filter the dataset to top 3 countries

@@

The data is collected for a 30-day period, during which users engaged in different formats of the tasks, namely `listen` (listen and translate into the source language), `reverse_tap` (order given tokens) and `reverse_translate` (read and translate into the source language). Only `listen` and `reverse_translate` tasks require typing thus they are more prone to errors^[Indeed, minimal edit distance is used to accept mistyping but it depends on a token, e.g. *you* will not be accepted for *your* even if edit distance is 1].

```{r, echo=FALSE, results='markup'}
es_en_md %>% 
  select(format, n_errors) %>%
  group_by(format) %>% 
  summarize(avg_error = mean(n_errors)) %>% 
  inner_join(fr_en_md %>%
              select(format, n_errors) %>%
              group_by(format) %>% 
              summarize(avg_error = mean(n_errors)), by="format") %>%
  knitr::kable(caption = "Average Error Rate",
               col.names = c("Task Type",
                             "Spanish",
                             "French"))
```

A task can contain from 1 to 14 tokens (depending on the language). All UD features were retrieved by @slam18 using the Google SyntaxNet dependency parser and the language-agnostic Universal Dependencies tagset^[Parse errors may occur.].

@@

In the `fr_en` dataset, `PUNCT` and `X` tokens have the largest error_rate. It was excluded from the graph as the former refers to `-` in constructions like `Qui sont-ils?` and the latter mostly to `t` in consruction like `Qu'a-t-il?`.  In both cases, error refers to the `-` character. The app does not penalise users for absence of punctuation marks, so they may just skip it thus "making a mistake". These mistakes mostly come from `reverse_tap` tasks, which does not assume any typing at all. It is a tagging or user-interface issue rather than a mistake that happend while learning a language^[These mistakes will be excluded from any analysis].

```{r, echo=FALSE, results='markup'}
fr_en_md %>% 
  select(session_id, format) %>% 
  right_join(fr_en_sessions %>% 
              filter(POS == 'PUNCT' & label == 1) %>% 
              select(session_id) %>% distinct(), by='session_id') %>% 
  select(format) %>% 
  table() %>% 
  knitr::kable(col.names = c('Format', 'Number of Errors'),
               caption='Distribution of Errors Tagged as PUNCT by Task Format')
```

The Spanish part of the graph does no include `SYM` and `X` tokens as the former occurs only once and the latter mostly is a `Sí` token and is rarely a mistake.

@@

## 3. NHST

@@

```{r}
fr_en_sessions %>%
  filter(POS=='ADP') %>% 
  separate_rows(morph, sep = "\\|") %>%
  separate(morph, into = c("feature", "value"), sep = "=")
```

>>>>

```{r}
library(tidyverse)
library(patchwork)
library(BayesFactor)
```

@@

``` {r, results='markup'}
cols_md = c('user_id', 'country', 'days', 'client',
            'session', 'format', 'time', 'session_id',
            'n_tokens', 'n_errors', 'prompt')
col_types_md = cols('f','f','d','f','f','f','i','f','i', 'i', 'f')

@@

```{r}
# filter the dataset to top 3 countries

@@

The data is collected for a 30-day period, during which users engaged in different formats of the tasks, namely `listen` (listen and translate into the source language), `reverse_tap` (drag tokens in a correct order) and `reverse_translate` (read and translate into the source language). Only `listen` and `reverse_translate` tasks require typing thus they are more prone to errors^[Indeed, minimal edit distance is used to handle mistyping but it depends on a token, e.g. *you* will not be accepted for *your* even if edit distance is only 1].

```{r, results='markup'}
es_en_md %>% 
  select(format, n_errors) %>%
  group_by(format) %>% 
  summarize(avg_error = mean(n_errors)) %>% 
  inner_join(fr_en_md %>%
              select(format, n_errors) %>%
              group_by(format) %>% 
              summarize(avg_error = mean(n_errors)), by="format") %>%
  knitr::kable(caption = "Average Error Rate",
               col.names = c("Task Type",
                             "Spanish",
                             "French"))
```

A task can contain from 1 to 14 tokens (depending on the language). Each token has a set of features assigned to it by @slam18 using the Google SyntaxNet dependency parser and the language-agnostic Universal Dependencies tagset^[Parse errors may occur.].

@@

In the `fr_en` dataset, `PUNCT` and `X` tokens have the largest error_rate. It was excluded from the graph as the former refers to `-` in constructions like `Qui sont-ils?` and the latter mostly to `t` in consruction like `Qu'a-t-il?`.  In both cases, error refers to the `-` character in a question. In general, the app does not penalise users for absence of punctuation marks, so they may just skip it thus "making a mistake".

```{r, results='markup'}
fr_en_md %>% 
  select(session_id, format) %>% 
  right_join(fr_en_sessions %>% 
              filter(POS == 'PUNCT' & label == 1) %>% 
              select(session_id) %>% distinct(), by='session_id') %>% 
  select(format) %>% 
  table() %>% 
  knitr::kable(col.names = c('Format', 'Number of Errors'),
               caption='Distribution of Errors Tagged as PUNCT by Task Format')
```

Another option is that this mistakes refers to word order issues. Some evidence for it comes from the fact these mistakes are mostly happen in `reverse_tap` tasks, which does not assume any typing at all. It is difficult to decide if it is a user-interface issue or a `real` mistake without seeing the actual user input thus no further exploration is possible.

The Spanish part of the graph does no include `SYM` and `X` tokens as the former occurs only once and the latter - a `Sí` token which is rarely put incorrectly.

@@

## 3. Exploring most common mistakes

**French-speaking learners**

The most "erroneous" token with the `ADP` tag is a preposition `de`. There are three variants of its spelling in Top 5 most common mistakes.

```{r, results='markup'}
fr_en_sessions %>%
  select(POS, token, label) %>% 
  filter(POS %in% c('ADP') & label == 1) %>% 
  group_by(token) %>%
  summarise(counter = n()) %>% 
  slice_max(counter, n=10) %>% 
  knitr::kable(caption = 'Most common mistakes for the ADP tag',
               col.names = c('Token', 'Quantity'))
```

It is quite unexpected, the two out of three variants refer to the the beginning of the sentence: `De` and `D'`. Apparently, it refers more to some labeling artifacts or a simple carelessness and does not deserve much attention. In these case, only `de`, `a`, `avec`, `en`, `chez`, `pour` will be explored more in-depth.

```{r}
extract_trigrams <- function(df, token_ids) {
  # return a matrix of trigrams with token_ids being in the centre 
  output <- cbind(
      as.matrix(df)[cbind(1:nrow(df), token_ids-1)],
      as.matrix(df)[cbind(1:nrow(df), token_ids)],
      as.matrix(df)[cbind(1:nrow(df), token_ids+1)])

    return(
      as.tibble(output) %>% 
        rename(previous = V1, token=V2, following=V3)
      )
}
```

```{r}
# extract trigrams from sentences where there are such these prepositions
sessions <- fr_en_sessions %>% 
  filter(POS == 'ADP' & token %in% c("a", 'de', 'avec', 'en', "chez", 'pour')) %>% 
  select(session_id, task_token_id, label)

# some sentences have two mistakes, but will look only the first one.
sessions <- sessions[!duplicated(sessions$session_id),]

by_pos <- fr_en_sessions %>% 
  select(session_id, task_token_id, POS) %>% 
  filter(session_id %in% sessions$session_id) %>% 
  pivot_wider(names_from = "task_token_id",
              values_from = "POS")

pos2labels <- sessions %>% 
  select(session_id, label) %>% 
  cbind(
    extract_trigrams(
      by_pos %>% 
        select(!session_id),
      sessions$task_token_id)
    )
```

```{r, fig.height=4, fig.width=10}
pos2labels %>% 
  select(label, previous) %>% 
  mutate(error = ifelse(label == 1, "Mistake", 'Correct')) %>% 
  ggplot(aes(y=previous))+
  geom_bar(aes(fill = error), position = position_stack(reverse = TRUE)) +
  scale_fill_manual("legend", values = c("Mistake" = "grey34", "Correct" = "grey"))+
  theme(legend.position = "top") + 
  labs(title = 'Proportion of mistakes for ADP token based on PoS of a previous token',
       x="", y="", caption = "ADP of this graph: a, de, avec, en, chez, pour")+
  theme_minimal()
```
There is a significant share of errors if a preposition goes after `ADJ`, `VERB` and `NOUN`. It might be useful to use features of these tokens to predict with logistic regression. All of them share the `Gender` features, so let's test if these feature correlates with errors.

1. Chi^2 test for `Gender` feature in `ADJ` indicates that that these two variables are independent as the p-value is much higher than 0.05. We can safely accept the null hypothesis.

```{r, results='markup'}
adj <- pos2labels %>% 
  filter(previous == 'ADJ') %>% 
  select(session_id) %>% 
  pull()

by_ud <- fr_en_sessions %>% 
  select(session_id, task_token_id, morph) %>% 
  filter(session_id %in% sessions$session_id) %>% 
  pivot_wider(names_from = "task_token_id",
              values_from = "morph")

ud2labels <- sessions %>% 
  select(session_id, label) %>% 
  cbind(
    extract_trigrams(
      by_ud %>% 
        select(!session_id),
     sessions$task_token_id))

ud2labels <- ud2labels %>% 
  select(1:3) %>% 
  filter(session_id %in% adj) %>% 
  select(session_id, label, previous) %>% 
  separate_rows(previous, sep = "\\|") %>% 
  separate(previous, into = c("feature", "value"), sep = "=")

features <- ud2labels %>% 
  pivot_wider(names_from = "feature",
              values_from = "value")

features %>% 
  drop_na(Gender) %>% 
  select(label, Gender) %>% 
  table() %>% 
  chisq.test()

features %>% 
  drop_na(Gender) %>% 
  select(label, Gender) %>% 
  table() %>% 
  contingencyTableBF(sampleType = "poisson")
```

To be on the safe side, I used the Bayesian framework for confirmation. The result is the same - the odds for the alternative hypothesis against the null are about 0.59:1.

2. The following are the results of the Chi^2 test for `Gender` feature in `VERB`

indicates that that these two variables are independent as the p-value is much higher than 0.05. We can safely accept the null hypothesis.

```{r, results='markup'}
verb <- pos2labels %>% 
  filter(previous == 'VERB') %>% 
  select(session_id) %>% 
  pull()

by_ud <- fr_en_sessions %>% 
  select(session_id, task_token_id, morph) %>% 
  filter(session_id %in% sessions$session_id) %>% 
  pivot_wider(names_from = "task_token_id",
              values_from = "morph")

ud2labels <- sessions %>% 
  select(session_id, label) %>% 
  cbind(
    extract_trigrams(
      by_ud %>% 
        select(!session_id),
     sessions$task_token_id))

ud2labels <- ud2labels %>% 
  select(1:3) %>% 
  filter(session_id %in% adj) %>% 
  select(session_id, label, previous) %>% 
  separate_rows(previous, sep = "\\|") %>% 
  separate(previous, into = c("feature", "value"), sep = "=")

features <- ud2labels %>% 
  pivot_wider(names_from = "feature",
              values_from = "value")

features %>% 
  drop_na(Gender) %>% 
  select(label, Gender) %>% 
  table() %>% 
  chisq.test()

features %>% 
  drop_na(Gender) %>% 
  select(label, Gender) %>% 
  table() %>% 
  contingencyTableBF(sampleType = "poisson")
```

@@
